Introduction Concurrent Programming with CXL

### 2020-05-15 book/paper.tex
Better example

\documentclass{book}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}

@@

\begin{figure}
\begin{code}
\verbatiminput{Count.cxl}
\end{code}
\caption{Counting up and down.}
\label{fig:updown}
\end{figure}

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:updown} presents a simple example.  The program defines two
methods \texttt{up()} and \texttt{down()}.
It also initializes three shared
variables: \texttt{count}, \texttt{upDone}, and \texttt{downDone}.
The \texttt{up()} method increments \texttt{count} and sets the \texttt{upDone}
flag.  The \texttt{down()} method is similar.
Next the program spawns two processes, one that invokes \texttt{up} and the
other \texttt{down}.  Afterward there are three processes---after
all, there was already one, and it spawned two new ones.  We will call
the one that was already there the \emph{main process}.

The main process waits for the two other processes to finish by checking
the \texttt{upDone} and \texttt{downDone} flags, simply in a loop.
Finally, it verifies that the value of \texttt{count} equals 0.  If not,
it reports the actual value of \texttt{count}.

\begin{quote}
What will happen?
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \texttt{count} will end up 0 after executing
\texttt{up} and \texttt{down} in parallel?
\item What are the possible values of \texttt{count} at the time of the
\texttt{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

What is going on is that the CXL program is compiled to machine instructions,
and it is the machine instructions that are executed by the underlying
machine.  The machine has instructions that load values from memory and store
values into memory, but it does not have instructions to atomically increment
or decrement values in memory locations.  So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved in unspecified and often random ways.
In fact, CXL will try \emph{all possible interleavings} of the processes
executing machine instructions.
And thus, the following is a possible interleaving of
the processes \texttt{up} and \texttt{down}:
\begin{enumerate}
\item \texttt{up} loads the value of \texttt{count}, which is 0;
\item \texttt{down} loads the value of \texttt{count}, which is still 0;
\item \texttt{down} subtracts one from the value that it loaded (0), and
stores $-1$ into \texttt{count};
\item \texttt{up} adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item \texttt{up} sets \texttt{upDone};
\item \texttt{down} sets \texttt{downDone}.
\end{enumerate}

The result in this particular interleaving is that \texttt{count} ends up
being 1.  Other interleavings can result in the value 0 or $-1$.  For the
CXL machine, those are all possible results.  When running CXL, it will
report the violations of the assertions.  It also provides an example
of an interleaving, like the one above, of how exactly the assertion
ended up failing.

@@

\begin{code}
\begin{verbatim}
    assert ((-1) <= count) and (count <= 1), count;
\end{verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
Count.cxl:1 def up():
    0 Jump 22
    1 Frame ('up', 'Count.cxl', 1, 5) None 10
Count.cxl:2     count = count + 1;
    2 Constant (1, 'Count.cxl', 2, 21)
    3 Load ('count', 'Count.cxl', 2, 13)
    4 Nary ('+', 'Count.cxl', 2, 19) 2
    5 Name ('count', 'Count.cxl', 2, 5)
    6 Store 1
Count.cxl:3     upDone = True;
    7 Constant (True, 'Count.cxl', 3, 14)
    8 Name ('upDone', 'Count.cxl', 3, 5)
    9 Store 1
   10 Return
   11 ...
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:updown}.}
\label{fig:updowncode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
Assertion failed ('assert', 'Count.cxl', 17, 1) 1

==== Safety violation ====
__main__ 0 [0,22-36]     37 { 'count':  0, 'upDone': False, 'downDone': False }
up       0 [1-5]          6 { 'count':  0, 'upDone': False, 'downDone': False }
down     0 [12-20]       21 { 'count': -1, 'upDone': False, 'downDone': True  }
up       0 [6-9]         10 { 'count':  1, 'upDone': True,  'downDone': True  }
__main__ 0 [37-41,43-48] 48 { 'count':  1, 'upDone': True,  'downDone': True  }
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:updown}.}
\label{fig:updownoutput}
\end{figure}

@@

Compiling the code in Figure~\ref{fig:updown} results in the CXL machine code
listed in Figure~\ref{fig:updowncode}.
The CXL machine is predominantly a \emph{stack machine}.
Most instructions pop values from the stack or push values onto the stack.
Initially there is one process that starts executing at instruction 0.
In this case, there is a \texttt{JUMP} instruction that sets the program
counter to 22.  At program counter 1 is the code for the \texttt{up} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

The \texttt{Frame} instruction has three arguments:
\begin{enumerate}
\item A token containing the name of the method;
\item A token containing the argument of the method (in this case the
method has no named argument);
\item A program counter of the method's \texttt{Return} instruction.
\end{enumerate}

A \emph{token} is a four-tuple containing a \emph{lexeme}, a file name, a
line number in the file, and a column number in the file.  For example,
\texttt{(’up’, ’Count.cxl’, 1, 5)} represents the lexeme \texttt{up} in
file \texttt{Count.xcl}, line 1, column 5.  Tokens make it easy to relate
the CXL machine instructions to the CXL source code.

The code generated from \texttt{count := count + 1} in line 2 of
\texttt{Count.cxl} is as follows:

@@

Figure~\ref{fig:updownoutput} shows the output produced by running the
\texttt{Count.cxl} program.
It starts by reporting that the assertion on line 17 failed and that
\texttt{count} has value 1 instead of 0:
\texttt{Assertion failed (’assert’, ’Count.cxl’, 17, 1) 1}.
Next it reports an execution that failed this assertion.  The output has
five columns:
\begin{enumerate}
\item The entry point of the process (\texttt{\_\_main\_\_} for the main process);
\item The tag of the process (in this case always 0);
\item The sequence of program counters of the CXL machine instructions that the process executed;
\item The current program counter of the process;
\item The contents of the shared memory.
\end{enumerate}

If we look at the middle three rows, we see that:
\begin{enumerate}
\item \texttt{up} executed instructions 1 through 5, loading the value of
\texttt{count} but stopping just before storing the value 1 into \texttt{count};
\item \texttt{down} executed instructions 12 through 20, which is all of the
\texttt{down} method, loading the value 0 from \texttt{count} and storing $-1$
into \texttt{count};
\item \texttt{up} continues execution, storing the value 1 into \texttt{count}.
\end{enumerate}

@@

>>>>

\documentclass{report}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}

@@

\begin{figure}
\begin{code}
\verbatiminput{Up.cxl}
\end{code}
\caption{Incrementing twice in parallel.}
\label{fig:inc}
\end{figure}

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.  The program defines a
method \texttt{up} that takes a parameter called \texttt{self}.
It also initializes two shared variables \texttt{count} and \texttt{done}.
Variable \texttt{done} is initialized with a dictionary that maps both 1 and 2
to \texttt{False}.
The \texttt{up()} method increments \texttt{count} and sets \texttt{done[self]}
to \texttt{True}.
Next the program spawns two processes, each invoking \texttt{up}.
Afterward there are three processes---after
all, there was already one, and it spawned two new ones.  We will call
the one that was already there the \emph{main process}.

The main process waits for the two other processes to finish by polling
both \texttt{done} flags.
Finally, it verifies that the value of \texttt{count} equals 2.  If not,
it reports the actual value of \texttt{count}.

\begin{quote}
What will happen?
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \texttt{count} will end up being 2?
\item What are the possible values of \texttt{count} at the time of the
\texttt{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

What is going on is that the CXL program is compiled to machine instructions,
and it is the machine instructions that are executed by the underlying
machine.  The details of this appear in the next chapter, but suffice it to
say that The machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved in unspecified and often random ways.
In fact, CXL will try \emph{all possible interleavings} of the processes
executing machine instructions.
And thus, the following is a possible interleaving of
the processes 1 and 2:
\begin{enumerate}
\item Process 1 loads the value of \texttt{count}, which is 0;
\item Process 2 loads the value of \texttt{count}, which is still 0;
\item Process 2 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 1 sets \texttt{upDone} to \texttt{True};
\item Process 2 sets \texttt{upDone} to \texttt{True}.
\end{enumerate}

The result in this particular interleaving is that \texttt{count} ends up
being 1.  When running CXL, it will
report the violations of the assertions.  It also provides an example
of an interleaving, like the one above, of how exactly the assertion
ended up failing.

@@

\begin{code}
\begin{verbatim}
    assert (count == 0) or (count == 1), count;
\end{verbatim}
\end{code}

@@

\begin{figure}
\begin{code}
\begin{verbatim}
Up.cxl:1 def up(self):
   0 Jump 12
   1 Frame ('up', 'Up.cxl', 1, 5) ('self', 'Up.cxl', 1, 8) 11
Up.cxl:2     count = count + 1;
   2 Constant (1, 'Up.cxl', 2, 21)
   3 Load ('count', 'Up.cxl', 2, 13)
   4 Nary ('+', 'Up.cxl', 2, 19) 2
   5 Name ('count', 'Up.cxl', 2, 5)
   6 Store 1
Up.cxl:3     done[self] = True;
   7 Constant (True, 'Up.cxl', 3, 18)
   8 LoadVar ('self', 'Up.cxl', 3, 10)
   9 Name ('done', 'Up.cxl', 3, 5)
   10 Store 2
   11 Return
Up.cxl:5 count = 0;
   12 Constant (0, 'Up.cxl', 5, 9)
   13 Name ('count', 'Up.cxl', 5, 1)
   14 Store 1
\end{verbatim}
\end{code}
\caption{The first part of the machine code corresponding to Figure~\ref{fig:inc}.}
\label{fig:inccode}
\end{figure}

\begin{figure}
\begin{code}
\begin{verbatim}
Assertion failed ('assert', '<stdin>', 12, 1) 1

==== Safety violation ====
__main__ 0 [0,12-29]     30 { 'count': 0, 'done': { 1: False, 2: False } }
up       1 [1-5]          6 { 'count': 0, 'done': { 1: False, 2: False } }
up       2 [1-10]        11 { 'count': 1, 'done': { 1: False, 2: True  } }
up       1 [6-10]        11 { 'count': 1, 'done': { 1: True,  2: True  } }
__main__ 0 [30-37,39-44] 44 { 'count': 1, 'done': { 1: True,  2: True  } }
\end{verbatim}
\end{code}
\caption{The output of running Figure~\ref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

Compiling the code in Figure~\ref{fig:inc} results in the CXL machine code
listed in Figure~\ref{fig:inccode}.
The CXL machine is predominantly a \emph{stack machine}.
Most instructions pop values from the stack or push values onto the stack.
Initially there is one process that starts executing at instruction 0.
In this case, there is a \texttt{JUMP} instruction that sets the program
counter to 22.  At program counter 1 is the code for the \texttt{up} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

The \texttt{Frame} instruction has three arguments:
\begin{enumerate}
\item A token containing the name of the method;
\item A token containing the name of the argument of the method;
\item A program counter of the method's \texttt{Return} instruction.
\end{enumerate}

A \emph{token} is a four-tuple containing a \emph{lexeme}, a file name, a
line number in the file, and a column number in the file.  For example,
\texttt{(’up’, ’Up.cxl’, 1, 5)} represents the lexeme \texttt{up} in
file \texttt{Up.xcl}, line 1, column 5.  Tokens make it easy to relate
the CXL machine instructions to the CXL source code.

The code generated from \texttt{count := count + 1} in line 2 of
\texttt{Up.cxl} is as follows:

@@

Figure~\ref{fig:incoutput} shows the output produced by running the
\texttt{Up.cxl} program.
It starts by reporting that the assertion on line 12 failed and that
\texttt{count} has value 1 instead of 0:
\texttt{Assertion failed (’assert’, ’Up.cxl’, 12, 1) 1}.
Next it reports an execution that failed this assertion.  The output has
five columns:
\begin{enumerate}
\item The entry point of the process (\texttt{\_\_main\_\_} for the main process);
\item The tag of the process (in this case always 0);
\item The sequence of program counters of the CXL machine instructions that the process executed;
\item The current program counter of the process;
\item The contents of the shared memory.
\end{enumerate}

If we look at the middle three rows, we see that:
\begin{enumerate}
\item Process 1 executed instructions 1 through 5, loading the value of
\texttt{count} but stopping just before storing 1 into \texttt{count};
\item Process 2 executed instructions 1 through 10, storing 1 into
into \texttt{count} and storing \texttt{True} into \texttt{done[2]};
\item Process 1 continues execution, storing value 1 into \texttt{count}
and storing \texttt{True} into \texttt{done[1]}.
\end{enumerate}

@@

\chapter{Naive Solutions}

Hopefully you have started thinking of how to solve the concurrency
problem and you may already have prototyped some solutions.
In this chapter we will go through a few reasonable but broken attempts.
At the heart of the problem is that we would like make sure that when
the \texttt{count} variable is being updated that no other process is
trying to do the same thing.  We call this a \emph{critical section}: a
set of instructions where only one process is allowed to execute at a
time.
